### 2013-10-11 Publications/2013-ISBI14/03-methods.tex
Final version submitted before new deadline (Oct 11)

\subsection{Digital Phantom}
Based on the fiber geometries of the digital phantom 
created for the \gls*{hardi} 
reconstruction Challenge held in ISBI 2013 
(San Francisco, US), we simulated high resolution 
(0.5mm isotropic) T1 (TE/TR=10/1500ms) and T2 
(TE/TR=90/5000ms) images, as well as two \gls*{dmri}
images (1.0mm isotropic, $b$=1200, 1 B0 image) with
32 and 64 evenly-distributed directions.
Diffusion is modeled by a restricted and a hindered
compartment, similar to \cite{assaf_composite_2005}.
The phantom includes \gls*{wm} fiber bundles,
\gls*{gm} and \gls*{csf}. Physical properties
(T1/T2 times in sec) used in simulation are
($0.832\pm0.010$ / $79\times10^{-3}\pm0.6\times10^{-3}$)
for \gls*{wm},
($1.331\pm0.013$ / $110\times10^{-3}\pm2.0\times10^{-3}$) for
\gls*{gm} and ($3.5\pm0.1$ / $0.25\pm0.01$) for \gls*{csf}.

@@

In order to create a realistic distortion, we
generated a synthetic phase-difference map 
consistent with the phantom, using the 
tools distributed with the \emph{FSL} package 
\cite{jenkinson_fsl_2012}  and
standard parameters ($\Delta$TE=2.46~ms. for the
field mapping and \emph{effective dwell time} of 
0.77~ms. for the \gls*{epi}).
We defined two regions of smooth dephasing and 
computed the corresponding \gls*{vsm}. Amplitude
of the dephasing maps can be modulated, enabling
us to evaluate the impact of the distortion magnitude.
We generated several \glspl*{vsm} with increasing
maximum shifts, ranging 3.80-7.60~mm, covering the
typical range of distortion observed in real datasets.

@@

The framework provides two different options for
\gls*{dwi} reconstruction and deterministic tractography.
Firstly, \emph{Diffusion Toolkit} \cite{wang_diffusion_2007}
for \gls*{dti}, is configured with 10 random
seeds per voxel by default. Secondly, \emph{MRTrix}
\cite{tournier_mrtrix:_2012} for \gls*{hardi}, with
default parameters set to use constrained spherical
deconvolution, maximum harmonic order of 6, and 150000
desired tracks.
For both options, the seeding regions can be set to use
either the distorted-corrected \gls*{wm} mask, or the 
regions used to generate the ground-truth. This second
seeding strategy mimics the usual procedure on real 
data, where regions are typically mapped from the
anatomically correct T1.

The evaluation framework is completed by automated 
assessment modules. We evaluated three characteristics
of the correction methods. 
Firstly, we assessed the geometrical correctness
reporting overlap indices of three tissue
probability maps (namely \gls*{csf}, \gls*{wm},
and \gls*{gm}), weighting the average by tissue
volumes.
Secondly, to evaluate the quality of the actual 
signal dropout correction, we studied the 
similarity volume by volume computing the $\ell_1$-norm
correlation index. We report this score
on the \textit{b0} and the average of the remaining \gls*{dwi}
volumes. Thirdly, we studied the impact on the
connectivity matrices reporting the number of
false positives (inexistent connections in the
gold-standard) and false negatives (or connections
lost).

>>>>

\subsection{Digital Phantom}
Based on the fiber geometries of the digital phantom 
created for the \emph{\Gls*{hardi} 
Reconstruction Challenge} held in ISBI 2013 
(San Francisco, US), we simulated high resolution 
(0.5mm isotropic) T1 (TE/TR=10/1500ms) and T2 
(TE/TR=90/5000ms) images, as well as two \gls*{dmri}
images (1.0mm isotropic, $b$=1200, 1 B0 image) with
32 and 64 evenly-distributed directions intended
for \gls*{dti} and \gls*{hardi} reconstructions, respectively.
Diffusion is modeled by a restricted and a hindered
compartment, similar to \cite{assaf_composite_2005}.
The phantom includes \gls*{wm} fiber bundles,
\gls*{gm} and \gls*{csf}. Physical properties
(T1/T2 times in sec) used in simulation are
($0.832\pm0.010$ / $79\times10^{-3}\pm0.6\times10^{-3}$)
for \gls*{wm},
($1.331\pm0.013$ / $110\times10^{-3}\pm2.0\times10^{-3}$) for
\gls*{gm} and ($3.5\pm0.1$ / $0.25\pm0.01$) for \gls*{csf}.

@@

In order to create a realistic distortion, we
generated a synthetic phase-difference map 
consistent with the phantom, using the 
tools distributed with the \emph{FSL} package 
\cite{jenkinson_fsl_2012}  and
standard parameters ($\Delta$TE=2.46~ms. for the
field mapping and \emph{effective dwell time} of 
0.77~ms. for the \gls*{epi}).
We defined two regions of smooth dephasing and 
computed the corresponding \gls*{vsm}. Amplitude
of the dephasing maps can be modulated, enabling
us to evaluate the impact of the distortion magnitude.
We generated several \glspl*{vsm} with increasing
maximum shifts (from 3.80 to 7.60~mm), covering 
the typical magnitudes of distortion observed 
in real datasets.

@@

The framework supports two different methods for
\gls*{dwi} reconstruction and deterministic tractography.
Firstly, \emph{Diffusion Toolkit} \cite{wang_diffusion_2007}
for \gls*{dti}, is configured with 10 random
seeds per voxel by default. Secondly, \emph{MRTrix}
\cite{tournier_mrtrix:_2012} for \gls*{hardi}, with
default parameters set to use constrained spherical
deconvolution, maximum harmonic order of 6, and 150000
desired tracks.
For both options, the seeding regions can be set to use
either the distorted-corrected \gls*{wm} mask, or the 
regions used to generate the ground-truth. This second
seeding strategy mimics the usual procedure on real 
data, where regions are typically mapped from the
anatomically correct T1.

The evaluation framework is completed by automated 
assessment modules. We evaluated three characteristics
of the correction methods. 
Firstly, we assessed the geometrical correctness
reporting overlap indices of three tissue
probability maps (namely \gls*{csf}, \gls*{wm},
and \gls*{gm}), weighting the average by tissue
volumes.
Secondly, to evaluate the quality of the actual 
signal dropout correction, we studied the 
similarity volume by volume computing the $\ell_1$-norm
correlation index. We report this score
on the \textit{b0} and the average of the remaining \gls*{dwi}
volumes. Thirdly, we studied the impact on the
connectivity matrices reporting the number of
false positives (inexistent connections in the
gold-standard) and false negatives (or missed connections).

### 2013-10-11 Publications/2013-ISBI14/04-results.tex
Final version submitted before new deadline (Oct 11)

All the modules described in the previous section were
conveniently integrated in workflows using \emph{nipype}
\cite{gorgolewski_nipype:_2011}. The choice of this tool
grants the reproducibility of the experiments,
and the evaluation workflow is publicly released.
The results of the proposed experiments are summarized
in \autoref{fig:results} and \autoref{table:results}.
The remaining of this section provides extended descriptions
and interpretation of the results.

@@

A summary of quantitative results, computed for the \gls*{dti} 
dataset with maximum shift of 3.80~mm,
is provided in \autoref{table:results}.
The best scores were obtained with the \gls*{reb} 
method, followed by \gls*{fmb}. 
The clear difference of accuracy between \gls*{reb} and \gls*{fmb}
with respect to \gls*{t2b} infers the latter may not be
an appropriate method for susceptibility correction. 
The described performance was constant for \gls*{fmb} 
and \gls*{reb} along the different magnitudes of 
distortion evaluated (maximum phase-encoding direction 
shifts in the range of 3.80-7.60~mm).

@@

Connectivity matrices derived from the \gls*{dti} dataset
were completely hindered by the high complexity of the 
fibers contained
in the phantom. Tractography discovered successfully only 4 out 
of 27 existing connections from the original (undistorted) data,
with more than 45 false connections.
With the \gls*{hardi} dataset (reconstructed with constrained 
spherical deconvolution) we found 23/27 connections,
but we still observed 40 false positives. Therefore, the 
experiments using the \gls*{dti} phantom were discarded. The
results presented in this subsection only refer to the 
\gls*{hardi} dataset.

@@

For the sake of completeness, \autoref{table:results-tractography}
reports the characteristic features of the tractography results
and connectivity matrices, for the fieldmap that caused a maximum
shift of 3.80~mm. Very similar results were obtained for
7.60~mm. This results, along with visual inspection
(\autoref{fig:results}, second and third rows), might point
to \gls*{fmb} as the best correction method.

@@

Even though all the surveyed methods produced visually 
sound results, our study suggested that \gls*{reb} is the 
most accurate method in terms of geometrical accuracy and
signal dropout recovery. 
The \gls*{t2b} method did not achieve the necessary high-standards
to recommend its use. Nonetheless, we understand that specific
methods with anisotropic regularization that completely
restrict deformations to the phase-encoding direction would perform
significantly better than the standard method presented.
Geometrical correctness of \gls*{dwi} data is fundamental 
in connectivity analysis to spatially locate the
regions which will define the nodes of the final connectivity
matrix.

>>>>

All the modules described in the previous section were
conveniently integrated in workflows using \emph{nipype}
\cite{gorgolewski_nipype:_2011}. The choice of this tool
grants the reproducibility of the experiments,
and the evaluation workflow is publicly released.
The results of this study are summarized
in \autoref{fig:results} and \autoref{table:results}.
The remaining of this section provides extended descriptions
and interpretation of these findings.

@@

A summary of quantitative results, computed for the \gls*{dti} 
dataset with maximum shift of 3.80~mm,
is provided in \autoref{table:results}.
The best scores were obtained with the \gls*{reb} 
method, followed by \gls*{fmb}. 
The clear difference of accuracy between \gls*{reb} and \gls*{fmb}
with respect to \gls*{t2b} infers the latter may not be
an appropriate method for susceptibility correction. 
The described performance was constant for \gls*{fmb} 
and \gls*{reb} along the different magnitudes of 
distortion evaluated (maximum shifts of 3.80-7.60~mm).

@@

Connectivity matrices derived from the \gls*{dti} dataset
were completely hindered by the high complexity of the 
fibers contained
in the phantom. Tractography discovered successfully only 4 out 
of 27 existing connections from the original (undistorted) data,
with more than 45 false connections.
With the \gls*{hardi} dataset we found 23/27 connections,
but we still observed 40 false positives. Therefore, the 
experiments using the \gls*{dti} phantom were discarded. The
results presented in this subsection only refer to the 
\gls*{hardi} dataset.

@@

For the sake of completeness, \autoref{table:results-tractography}
reports the characteristic features of the tractography results
and connectivity matrices, for the fieldmap that caused a maximum
shift of 3.80~mm. A very similar outcome was obtained for
7.60~mm. These results, along with visual inspection
(\autoref{fig:results}, second and third rows), might point
to \gls*{fmb} as the best correction method.

@@

Even though all the surveyed methods produced visually 
sound results, our study suggested that \gls*{reb} is the 
most accurate method in terms of geometrical agreement and
signal dropout recovery. 
The \gls*{t2b} method did not achieve the necessary high-standards
to recommend its use. Nonetheless, we understand that specific
methods with anisotropic regularization that completely
restrict deformations to the phase-encoding direction would perform
significantly better than the standard method presented.
Geometrical correctness of \gls*{dwi} data is fundamental 
in connectivity analysis to spatially locate the
regions which will define the nodes of the final connectivity
matrix.

### 2013-10-11 Publications/2013-ISBI14/05-conclusions.tex
Final version submitted before new deadline (Oct 11)

This paper proposes an evaluation framework to
comprehensively analyze the impact of susceptibility
induced distortion on tractography from \gls*{dmri}
data. Inaccuracy on tractography produce errors
and increased variability in subsequent connectivity 
analyses of the whole-brain. This effect is suspected
to be larger in regions influenced by large distortions.
Even though, this study found out that the connectivity
matrix is more largely affected by the tractography method.
We publicly release the framework and also contribute
with the evaluation of the most widely-used and
readily-available correction methodologies. 
The reverse-encoding method (\gls*{reb}) achieved the
best results correcting for susceptibility-induced 
artifacts. We also reported visual evaluation of 
tractography and quantitative results for the 
extracted connectivity matrices.

>>>>

This paper proposes an evaluation framework to
comprehensively analyze the impact of susceptibility
induced distortion on connectivity analyses from \gls*{dmri}
data. Inaccuracy on tractography produce errors
and increased variability in the extracted connectivity 
matrices of the whole-brain. This effect is suspected
to influence those regions with larger distortions.
However, this study found out that the connectivity
matrix is more largely affected by the tractography method.
We publicly release the framework and also contribute
with the evaluation of the most widely-used and
readily-available correction methodologies. 
The reverse encoding-based method achieved the
best results correcting for susceptibility-induced 
artifacts. We also reported visual evaluation of 
tractography and quantitative results for the 
extracted connectivity matrices.
